NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-05 Manuscript/manuscriptPS.tex
Fixed few typos; asked for a few clarifications.

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H NMR experiments~\cite{browning80}.
Slices of the R-PDFL spectra and the resulting order parameters values
are shown in the supplementary information (Fig. \ref{DPslices}). 
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice clearly shows that the order parameter of
the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
suggests that the larger order parameter of the $\alpha$-carbon 
is positive and the deviation towards negative values with longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available.

@@

Calcium binding affinity to membranes containing the negative charged PS lipids can be
experimentally measured by detecting the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the dehydrated lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding affinity.
The only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameters, 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions
along membrane normal, where almost all Ca$^{2+}$ ions bind to the membrane interface in
all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

The headgroup order parameters of POPS experimentally measured from a POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}).
In experiments, the order parameter of the POPS $\beta$-carbon increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 with underestimated binding affinity.
In addition, different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly describe interactions between the
PS headgroup and calcium ions in MD simulations.

>>>>

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }\todo{
  1) Use diamonds for DOPS, spheres for POPS.
  2) Error bars?
  3) Change the y-label to $S_\mathrm{CH}$, and invert the y-axis, as in Fig. 3.
  4) Make lower g1 POPC visible, e.g., by slightly larger point.
  }
\end{figure}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}). 
Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of the PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice clearly shows that the order parameter of
the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
Could we explain in a few words how this is clearly seen?}
which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
suggests that the larger order parameter of the $\alpha$-carbon 
is positive and the deviation towards negative values with longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
confirming \todo{Or rather 'corroborating'?}
the interpretation that the smaller order parameter is negative.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available.

@@

Calcium binding affinity to membranes containing negative charged PS lipids can be
measured by detecteding the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the dehydrated lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of ion free reference state
in the presence of negatively charged lipids, the simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as a function of Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding binding affinity.
Only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the changes in order parameter 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
in all the other tested models. This is evident in the calcium density distributions
along membrane normal where almost all Ca$^{2+}$ ions bind to the membrane interface in
all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}).
The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
tested simulation models, including the CHARMM36 with underestimated binding affinity.
In addition, different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with the added calcium.
For example, both order parameters decrease in Berger simulations but increase
in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
more complicated. This is in contrast to the PC headgroup, where
qualitatively correct reponse to the bound ions is observed
in all simulation models despite of the significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
the improvement of force fields is necessary to correctly describe interactions between
PS headgroup and calcium ions in MD simulations.
